The Concurrent Calculi Formalisation Benchmark

### 2024-01-31 short-paper/binders.tex
Edits and comments

The encoding of binding constructs has been extensively studied for type systems based on lambda calculi, but process calculi present unique challenges.
$\pi$-calculi typically include several different binding constructs: inputs bind a received name or value, recursive processes bind recursion variables, and restrictions bind names.
The first two act similarly to the binders known from lambda calculi, but restrictions are more challenging due to scope extrusion.

The case study \cite{AmbalLS21} compares four approaches to encoding binders in Coq as a first step towards developing tools for working with higher-order process calculi.
They found that working directly with de Bruijn indices was easiest and shortest since the existing approaches developed for lambda calculus binders do not work well with scope extrusion.
Specifically, the locally nameless representation~\cite{Chargueraud2012} cannot avoid direct manipulation of de Bruijn indices when defining the semantics of scope extrusion, cofinite quantification does not provide any benefits when working under binders, nominal approaches~\cite{Pitts2003} require explicitly giving and validating sets of free names during scope extrusion, and higher-order abstract syntax (HOAS)~\cite{Pfenning1988} requires additional axioms to work with contexts.

Additionally, many proof assistants lack support for the proof principles required to easily work with scope extrusion, i.e.\ induction principles on functions to work with HOAS encodings.
Some proof assistants such as Abella~\cite{Baelde2014} support HOAS natively, but working with open terms as required to encode scope extrusion in these systems is generally very challenging~\cite{Momigliano2012}.

% Start of related work
\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.

>>>>

The encoding of binding constructs has been extensively studied for type systems based on $\lambda$-calculi, but process calculi present unique challenges.
$\pi$-calculi typically include several different binding constructs: inputs bind a received name or value, recursive processes bind recursion variables, and restrictions bind names.
The first two act similarly to the binders known from $\lambda$-calculi, but restrictions are more challenging due to scope extrusion.

The case study \cite{AmbalLS21} compares four approaches to encoding binders in Coq as a first step towards developing tools for working with higher-order process calculi.
They found that working directly with de Bruijn indices was easiest and shortest since the existing approaches developed for $\lambda$-calculus binders do not work well with scope extrusion.
Specifically, the locally nameless representation~\cite{Chargueraud2012} cannot avoid direct manipulation of de Bruijn indices when defining the semantics of scope extrusion; cofinite quantification provides no benefits when working under binders; nominal approaches~\cite{Pitts2003} require explicitly giving and validating sets of free names during scope extrusion; and higher-order abstract syntax (HOAS)~\cite{Pfenning1988} requires additional axioms to work with contexts.

Additionally, many proof assistants lack support for the proof principles required to easily work with scope extrusion, \ie, induction principles on functions to work with HOAS encodings.
Some proof assistants such as Abella~\cite{Baelde2014} support HOAS natively, but working with open terms as required to encode scope extrusion in these systems is generally very challenging~\cite{Momigliano2012}.

% Start of related work
\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion. \dz{Not sure if relevant to mention.}

### 2024-01-31 short-paper/lin.tex
Edits and comments

In mechanized meta-theory addressing linearity means choosing an
appropriate representation of the linear context. The crux of the matter
is \emph{context splitting} as showcased by the typing rule for parallel composition.
% the introduction rule of the tensor.
While a linear context is best seen as a multi-set, most proof
assistants have better support for list processing.  Although the
latter representation is intuitive, it is overly concrete and may
require establishing dozens and dozens of technical lemmas that are
foreign to the mathematics of the problem under study, say proving
type preservation for session types. In this space, several designs
are possible: one can label occurrences of resources constraining
their usage, e.g.~\cite{CicconeP20}, or \alb{something --- ask
  Francisco} as in \cite{Castro2020}. Some authors instead impose a
multiset structure over lists, see~\cite{ChaudhuriLR19,Danielsson12}:
still, the effort in developing the infrastructure is significant.

A different approach is to bypass the problem of context splitting by
adopting
% to  meta-theoretic properties
familiar ideas from algorithmic linear type checking: this is known as
 \emph{typing with leftovers}, as exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}. Whatever the choice, list-based
encodings can be refined to be intrinsically-typed, if the proof
assistant does support dependent types, see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}.

For something completely different, one could adopt a
\emph{sub-structural} meta-logical framework: here it is the framework
itself that handles resource distribution and the users just have to map
their linear operations to the ones offered by the framework. In fact,
the linear function space will handle all context operations
implicitly, including splitting and substitution. One such a framework
is \emph{Celf}~\cite{Schack-Nielsen:IJCAR08}, see the encoding of
session types in~\cite{Bock2016}. Unfortunately, \emph{Celf} does not
yet fully support the verification of meta-theoretic properties. A compromise is the so called
\emph{two-level} approach, where one encodes a sub-structural
specification logic in a mainstream proof assistant and then use that
logic to state and prove linear properties, see for a recent
example~\cite{Felty:MSCS21}.

>>>>

In mechanised meta-theory, addressing linearity means choosing an
appropriate representation of the linear context. The crux of the matter
is \emph{context splitting}, as showcased by the typing rule for parallel composition.
% the introduction rule of the tensor.
While a linear context is perhaps best seen as a multi-set, most proof
assistants have better support for list processing.  The
latter representation is intuitive, but may require establishing a 
large number of technical lemmata that are
orthogonal to the mathematics of the problem under study, say proving
type preservation for session types. In this space, several designs
are possible: one can label occurrences of resources to constrain
their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}. Some authors instead impose a
multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12});
still, the effort required to develop the infrastructure is significant.

A different approach is to bypass the problem of context splitting by
adopting
% to  meta-theoretic properties
familiar ideas from algorithmic linear type checking: this is known as
 \emph{typing with leftovers}, as exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}. Whatever the choice, list-based
encodings can be refined to be intrinsically-typed, if the proof
assistant does support dependent types (see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).

For something completely different, one could adopt a
\emph{sub-structural} meta-logical framework: here the framework
itself handles resource distribution, and users need only map
their linear operations to the ones offered by the framework. In fact,
the linear function space will handle all context operations
implicitly, including splitting and substitution. One such framework
is \emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of
session types in~\cite{Bock2016}). Unfortunately, \emph{Celf} does not
yet fully support the verification of meta-theoretic properties. A compromise is the so called
\emph{two-level} approach, where one encodes a sub-structural
specification logic in a mainstream proof assistant and then uses that
logic to state and prove linear properties (for a recent
example, see~\cite{Felty:MSCS21}).

### 2024-01-31 short-paper/short-paper.tex
first draft of related work abut linearity
Add first draft piece about binders.
Edits and comments

\begin{abstract}
  The POPLMark challenge and its follow-up POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. However, both POPLMark and
  POPLMark Reloaded were purposely limited in scope, with benchmark problems
  that do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems with a
  focus on the challenges and issues that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{wondering whether it would make sense to say ``\dots using formal models of concurrent and distributed programming languages in the form of process calculi'' or something similar to stress that process calculi are not the subject of the challenges, but ``only'' an appropriate tool for this kind of endeavor}  Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

Several papers in this realm already include mechanised proofs
(e.g.~\cite{Castro-Perez2021,Castro2020,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques
(e.g.~\cite{Bengtson2016, Gay2001,
  Thiemann2019, DBLP:conf/forte/ZalakainD21}).  Our
experience, however, is that choosing the appropriate proof mechanisation techniques and 
tools remains a significant challenge with a limited understanding of their 
respective tradeoffs.  This often leads researchers toward a trial-and-error approach, 
resulting in sub-optimal and ad-hoc solutions. In turn, this increases the overall effort 
required for proof mechanisation, and hampers the reuse of techniques and results.
%
It is therefore our view that the research community in message-passing
concurrency (and process calculi, more specifically) will benefit from a standardised set
of challenge problems.

@@

\section{Motivation}\label{sec:motivation}

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.  Still, these (and other) applications need the basic techniques that our challenge problems emphasize:
as per design goal \ref{item:goal-reusability},
our problems are drawn from the basic metatheory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

@@

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions.
It is not obvious that the techniques used to solve the original challenges about binders can be easily extended to reasoning about scope extrusion.
For example, techniques like higher-order abstract syntax, or well-scoped
De\ Bruijn indices do not have the concept of free variables and thus encoding this challenge with them requires some attention.

@@

@@

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not cover here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.

>>>>

\begin{abstract}
  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{wondering whether it would make sense to say ``\dots using formal models of concurrent and distributed programming languages in the form of process calculi'' or something similar to stress that process calculi are not the subject of the challenges, but ``only'' an appropriate tool for this kind of endeavor} \dz{Maybe something like, ``\ldots that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.''?} Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

Several papers in this realm already include mechanised proofs
(\eg~\cite{Castro-Perez2021,Castro2020,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques
(\eg~\cite{Bengtson2016, Gay2001,
  Thiemann2019, DBLP:conf/forte/ZalakainD21}).  Our
experience, however, is that choosing the appropriate proof mechanisation techniques and 
tools remains a significant challenge with a limited understanding of their 
respective tradeoffs.  This often leads researchers toward a trial-and-error approach, 
resulting in sub-optimal and ad-hoc solutions. In turn, this increases the overall effort 
required for proof mechanisation, and hampers the reuse of techniques and results.
%
It is therefore our view that the research community in message-passing
concurrency (and process calculi, more specifically) will benefit from a standardised set
of challenge problems.

@@

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.  Still, these (and other) applications need the basic techniques that our challenge problems emphasize:
as per design goal \ref{item:goal-reusability},
our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

@@

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions.
It is not obvious that the techniques used to solve the original challenges about binders can be easily extended to reasoning about scope extrusion.
For example, techniques like higher-order abstract syntax, or well-scoped
de\ Bruijn indices do not have the concept of free variables and thus encoding this challenge with them requires some attention.

@@

\section{Possible approaches}\label{sec:approaches}

\alb{Some introductory text to lead to the literature review. \sout{We may want to move this after the next Section, if we want to mirror the 3 challnges}}
\dz{Agreed, I moved it for now.} \\
\dz{Should we not also include a subsection on coinduction?} \\
\dz{I changed the name of this section to ``Possible approaches'' from ``Motivation''. Thoughts?}

\subsubsection{Linearity.}
\input{lin}

\subsubsection{Binders.}
\input{binders}

@@

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not cover here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.
